NumPy---Array Programming in Python

### 2020-01-20 body.tex
Minor edits

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It consists of an $n$-dimensional array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for array programming.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.9\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated meta-data fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{e,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
   }
  \label{fig:array-concepts}
\end{figure*}

The NumPy array is a data structure that efficiently stores and accesses
$n$-dimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
enables a wide variety of scientific computation.
It consists of a pointer to memory, along with meta-data used to interpret the
data stored there, notably {\em data type}, {\em shape}, and {\em strides}
(Fig.~\ref{fig:array-concepts}a).

The {\em data type} describes the nature of elements stored in an array.
An array has a single data type, and each array element occupies the same
number of bytes in memory.
Examples of data types include real and complex numbers (of lower and higher
precision), strings, timestamps, and pointers to Python objects.

@@

{\em Strides} are necessary to interpret computer memory, which stores elements
linearly, as $n$-dimensional arrays.
It describes the number of bytes to move forward in memory to jump from row to
row, column to column, and so forth.
Consider, for example, a 2-D array of floating point numbers with shape
$(4, 3)$, where each element occupies 8 bytes in memory.
To move between consecutive columns we need to jump forward 8 bytes in memory,
and to access the next row $3 \times 8 = 24$ bytes.
The strides of that array are therefore $(24, 8)$.  NumPy is able to
store arrays in either C or Fortran memory order, iterating
first over either rows or columns.  This allows external libraries
written in those languages to directly access NumPy array data in memory.

To manipulate the data structure in an intuitive manner, we provide users with
a simple yet powerful array expression syntax.
It supports operations such as indexing (selecting subarrays) and element-wise
calculations.
A high level syntax allows users to express themselves succinctly, while NumPy
deals with the underlying mechanics of making operations fast.

@@

To complement the array syntax, NumPy includes several functions that perform
element-wise calculations on arrays, including arithmetic, statistics, and
trigonometry.
They aim to loop over array elements near-optimally, taking into consideration,
e.g., strides, in order to best utilize the computer's fast cache memory.
Vectorized operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.

Many of these functions, such as summation, support reductions: aggregating
results across one or more dimensions of the array.
For example, summing a $d$-dimensional array over $c$ axes results in a
$(d-c)$-dimensional array (Fig.~\ref{fig:array-concepts}e).

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
  \caption{\textbf{Proliferation of Arrays.} The NumPy array is a classic and simple in-memory representation of array data. Since its inception, numerous other array representations have been developed, including distributed arrays---which are stored on multiple computers, GPU arrays---which utilize graphics processing units for fast computation, computation graphs---which postpone and combine calculations before executing them efficiently, and sparse arrays---which typically contain few non-zero values, and store only those in memory.}
  \label{fig:array-proliferation}
\end{figure*}

@@

Given a rapid proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.

>>>>

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It consists of an multidimensional array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for array programming.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{e,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
   }
  \label{fig:array-concepts}
\end{figure*}

The NumPy array is a data structure that efficiently stores and accesses
multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
enables a wide variety of scientific computation.
It consists of a pointer to memory, along with metadata used to interpret the
data stored there, notably {\em data type}, {\em shape}, and {\em strides}
(Fig.~\ref{fig:array-concepts}a).

The {\em data type} describes the nature of elements stored in an array.
n array has a single data type, and each array element occupies the same
number of bytes in memory.
Examples of data types include real and complex numbers (of lower and higher
precision), strings, timestamps, and pointers to Python objects.

@@

{\em Strides} are necessary to interpret computer memory, which stores elements
linearly, as multidimensional arrays.
It describes the number of bytes to move forward in memory to jump from row to
row, column to column, and so forth.
Consider, for example, a 2-D array of floating point numbers with shape
$(4, 3)$, where each element occupies 8 bytes in memory.
To move between consecutive columns we need to jump forward 8 bytes in memory,
and to access the next row $3 \times 8 = 24$ bytes.
The strides of that array are therefore $(24, 8)$.  NumPy is able to
store arrays in either C or Fortran memory order, iterating
first over either rows or columns.  This allows external libraries
written in those languages to directly access NumPy array data in memory.

To manipulate the data structure in an intuitive manner, we provide users with
a simple yet powerful array expression syntax.
It supports operations such as indexing (selecting subarrays) and elementwise
calculations.
A high level syntax allows users to express themselves succinctly, while NumPy
deals with the underlying mechanics of making operations fast.

@@

To complement the array syntax, NumPy includes functions that perform
elementwise calculations on arrays, including arithmetic, statistics, and
trigonometry.
They aim to loop over array elements near-optimally, taking into consideration,
e.g., strides, in order to best utilize the computer's fast cache memory.
Vectorized operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.

Many of these functions, such as summation, support reductions: aggregating
results across one or more dimensions of the array.
For example, summing a $n$-dimensional array over $d$ axes results in a
$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}e).

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.6\textwidth]{static/sketches/duck-arrays}
  \caption{\textbf{Proliferation of Arrays.} The NumPy array is a classic and simple in-memory representation of array data. Since its inception, numerous other array representations have been developed, including distributed arrays---which are stored on multiple computers, GPU arrays---which utilize graphics processing units for fast computation, computation graphs---which postpone and combine calculations before executing them efficiently, and sparse arrays---which typically contain few non-zero values, and store only those in memory.}
  \label{fig:array-proliferation}
\end{figure*}

@@

Given the proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.

### 2020-01-20 methods.tex
Minor edits

The \emph{universal functions}, or more concisely, \emph{ufuncs},
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two values and computes
$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
the ufunc will take care of looping over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-d array with length 3, and
\code{y} is a 2-d array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$.  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to complete the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.

>>>>

The \emph{universal functions}, or more concisely, \emph{ufuncs},
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two values and computes
$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
the ufunc will take care of looping over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-D array with length 3, and
\code{y} is a 2-D array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$.  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to complete the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.
